---
title: "analysis_train"
author: "Michael Hermsen"
date: "2023-05-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Set up

This script will be used to make figures/results for the age/CSC project.

Variables should be:

* encounter
* patient
* lfspprt_episode
* sofa_total_48hr
* age_years
* race
* covid
* died
* vent_ever

In this script, the data is assumed to be 1 row per life support episode, with
the 48-hour maximum SOFA score calculated at the beginning of the LSE.

```{r load_libs, message=FALSE}
library(ggplot2)
library(pROC)
library(dplyr)
library(rms)
library(gtsummary)
library(splines)
library(caret)
library(webshot)
library(tidyr)
library(CalibrationCurves)
```


```{r load_data, include=FALSE}

# Set your path to training data here
setwd("C:/Users/t.cri.mhermsen/Documents/UCMC_data_clean/Run_UC_NW_train")

# make sure to load UC data first
data <- read.csv("data_for_analysis.csv")  %>%
  arrange(encounter, lfspprt_episode)

# flag data as UC
data$UC <- 1

# set path to NW data here, if different
# setwd()

# now load NW data
# read.csv()

# rbind UC + NW data together, save as "data"
# data <- data %>% rbind()

# flag NW as testing set
data$UC <- ifelse(is.na(data$UC), 0, data$UC)

```

## Table 1

We will create several "Table 1s" in this script. Some will be generated based on
all data (UC + NW combined), and others will be generated based on only UC or NW
data. Additionally, we will have tables generated at the patient, LSE, and encounter level.

``` {r tab1_sofagroups, echo=FALSE}
# create SOFA groups variable for table 1s
data <- data %>%
  group_by(encounter, lfspprt_episode) %>%
  mutate(sofa_group = cut(sofa_total_48hr, breaks=c(-Inf, 7, 11, Inf),
                          labels= c("SOFA 0-7",
                                    "SOFA 8-11",
                                    "SOFA > 11")))

```

Patient Level Table 1 (UC and NW Data)
```{r table1_UCNW_pt, echo=FALSE}

tbl1_UCNW_pt <- data %>%
  ungroup %>%
  arrange(patient, encounter, lfspprt_episode) %>%
  distinct(patient, .keep_all=TRUE) %>% # keeps first instance
  select(age_years,
         sex,
         race,
         sofa_total_48hr,
         sofa_group,
         vent_ever,
         died) %>%
  tbl_summary(
    label = list(age_years = "Age",
                 sofa_total_48hr = "48-hour Maximum SOFA score at beginning of LSE",
                 sofa_group = "SOFA distribution",
                 vent_ever = "Ever on mechanical ventilation",
                 race = "Race/ethnic group",
                 sex = "Sex",
                 died = "Death or discharge to hospice"),
    statistic = age_years ~ "{mean} ({sd})"
  )

tbl1_UCNW_pt

save(tbl1_UCNW_pt, file="table1_UCNW_pt.RData")

gt::gtsave(as_gt(tbl1_UCNW_pt), filename = "table1_UCNW_pt.png")

```
Patient Level Table 1 (UC data only)
``` {r table1_UC_pt, echo=FALSE}

data_UC <- data %>%
  filter(UC == 1)


tbl1_UC_pt <- data_UC %>%
  ungroup %>%
  arrange(patient, encounter, lfspprt_episode) %>%
  distinct(patient, .keep_all=TRUE) %>% # keeps first instance
  select(age_years,
         sex,
         race,
         sofa_total_48hr,
         sofa_group,
         vent_ever,
         died) %>%
  tbl_summary(
    label = list(age_years = "Age",
                 sofa_total_48hr = "48-hour Maximum SOFA score at beginning of LSE",
                 sofa_group = "SOFA distribution",
                 vent_ever = "Ever on mechanical ventilation",
                 race = "Race/ethnic group",
                 sex = "Sex",
                 died = "Death or discharge to hospice"),
    statistic = age_years ~ "{mean} ({sd})"
  )

tbl1_UC_pt

save(tbl1_UC_pt, file="table1_UC_pt.RData")

gt::gtsave(as_gt(tbl1_UC_pt), filename = "table1_UC_pt.png")

```

Patient Level Table 1 (NW data only)
``` {r table1_NW_pt, echo=FALSE, eval=FALSE}

data_NW <- data %>%
  filter(UC == 0)


tbl1_NW_pt <- data_NW %>%
  ungroup %>%
  arrange(patient, encounter, lfspprt_episode) %>%
  distinct(patient, .keep_all=TRUE) %>% # keeps first instance
  select(age_years,
         sex,
         race,
         sofa_total_48hr,
         sofa_group,
         vent_ever,
         died) %>%
  tbl_summary(
    label = list(age_years = "Age",
                 sofa_total_48hr = "48-hour Maximum SOFA score at beginning of LSE",
                 sofa_group = "SOFA distribution",
                 vent_ever = "Ever on mechanical ventilation",
                 race = "Race/ethnic group",
                 sex = "Sex",
                 died = "Death or discharge to hospice"),
    statistic = age_years ~ "{mean} ({sd})"
  )

tbl1_NW_pt

save(tbl1_NW_pt, file="table1_NW_pt.RData")

gt::gtsave(as_gt(tbl1_NW_pt), filename = "table1_NW_pt.png")

```

Encounter Level Table 1 (UC and NW data)
``` {r table1_UCNW_enc, echo=FALSE}
tbl1_UCNW_enc <- data %>%
  ungroup %>%
  arrange(encounter, lfspprt_episode) %>%
  distinct(encounter, .keep_all=TRUE) %>% # keeps first instance
  select(age_years,
         sex,
         race,
         sofa_total_48hr,
         sofa_group,
         vent_ever,
         died) %>%
  tbl_summary(
    label = list(age_years = "Age",
                 sofa_total_48hr = "48-hour Maximum SOFA score at beginning of LSE",
                 sofa_group = "SOFA distribution",
                 vent_ever = "Ever on mechanical ventilation",
                 race = "Race/ethnic group",
                 sex = "Sex",
                 died = "Death or discharge to hospice"),
    statistic = age_years ~ "{mean} ({sd})"
  )

tbl1_UCNW_enc

save(tbl1_UCNW_enc, file="table1_UCNW_enc.RData")

gt::gtsave(as_gt(tbl1_UCNW_enc), filename = "table1_UCNW_enc.png")
```

Encounter Level Table 1 (UC data only)
``` {r table1_UC_enc, echo=FALSE}

tbl1_UC_enc <- data_UC %>%
  ungroup %>%
  arrange(encounter, lfspprt_episode) %>%
  distinct(encounter, .keep_all=TRUE) %>% # keeps first instance
  select(age_years,
         sex,
         race,
         sofa_total_48hr,
         sofa_group,
         vent_ever,
         died) %>%
  tbl_summary(
    label = list(age_years = "Age",
                 sofa_total_48hr = "48-hour Maximum SOFA score at beginning of LSE",
                 sofa_group = "SOFA distribution",
                 vent_ever = "Ever on mechanical ventilation",
                 race = "Race/ethnic group",
                 sex = "Sex",
                 died = "Death or discharge to hospice"),
    statistic = age_years ~ "{mean} ({sd})"
  )

tbl1_UC_enc

save(tbl1_UC_enc, file="table1_UC_enc.RData")

gt::gtsave(as_gt(tbl1_UC_enc), filename = "table1_UC_enc.png")
```

Encounter Level Table 1 (NW data only)
``` {r table1_NW_enc, echo=FALSE, eval=FALSE}

tbl1_NW_enc <- data_NW %>%
  ungroup %>%
  arrange(encounter, lfspprt_episode) %>%
  distinct(encounter, .keep_all=TRUE) %>% # keeps first instance
  select(age_years,
         sex,
         race,
         sofa_total_48hr,
         sofa_group,
         vent_ever,
         died) %>%
  tbl_summary(
    label = list(age_years = "Age",
                 sofa_total_48hr = "48-hour Maximum SOFA score at beginning of LSE",
                 sofa_group = "SOFA distribution",
                 vent_ever = "Ever on mechanical ventilation",
                 race = "Race/ethnic group",
                 sex = "Sex",
                 died = "Death or discharge to hospice"),
    statistic = age_years ~ "{mean} ({sd})"
  )

tbl1_NW_enc

save(tbl1_NW_enc, file="table1_NW_enc.RData")

gt::gtsave(as_gt(tbl1_NW_enc), filename = "table1_NW_enc.png")
```

LSE Level Table 1 (UC and NW data)
``` {r table1_UCNW_lse, echo=FALSE}

tbl1_UCNW_lse <- data %>%
  ungroup %>%
  arrange(encounter, lfspprt_episode) %>% # no need to restrict rows... each row is already distinct LSE
  select(age_years,
         sex,
         race,
         sofa_total_48hr,
         sofa_group,
         vent_ever,
         died) %>%
  tbl_summary(
    label = list(age_years = "Age",
                 sofa_total_48hr = "48-hour Maximum SOFA score at beginning of LSE",
                 sofa_group = "SOFA distribution",
                 vent_ever = "Ever on mechanical ventilation",
                 race = "Race/ethnic group",
                 sex = "Sex",
                 died = "Death or discharge to hospice"),
    statistic = age_years ~ "{mean} ({sd})"
  )

tbl1_UCNW_lse

save(tbl1_UCNW_lse, file="table1_UCNW_lse.RData")

gt::gtsave(as_gt(tbl1_UCNW_lse), filename = "table1_UCNW_lse.png")
```

LSE Level Table 1 (UC data only)
``` {r table1_UC_lse, echo=FALSE}

tbl1_UC_lse <- data_UC %>%
  ungroup %>%
  arrange(encounter, lfspprt_episode) %>% # no need to restrict rows... each row is already distinct LSE
  select(age_years,
         sex,
         race,
         sofa_total_48hr,
         sofa_group,
         vent_ever,
         died) %>%
  tbl_summary(
    label = list(age_years = "Age",
                 sofa_total_48hr = "48-hour Maximum SOFA score at beginning of LSE",
                 sofa_group = "SOFA distribution",
                 vent_ever = "Ever on mechanical ventilation",
                 race = "Race/ethnic group",
                 sex = "Sex",
                 died = "Death or discharge to hospice"),
    statistic = age_years ~ "{mean} ({sd})"
  )

tbl1_UC_lse

save(tbl1_UC_lse, file="table1_UC_lse.RData")

gt::gtsave(as_gt(tbl1_UC_lse), filename = "table1_UC_lse.png")
```

LSE Level Table 1 (NW data only)
``` {r table1_NW_lse, echo=FALSE, eval=FALSE}

tbl1_NW_lse <- data_NW %>%
  ungroup %>%
  arrange(encounter, lfspprt_episode) %>% # no need to restrict rows... each row is already distinct LSE
  select(age_years,
         sex,
         race,
         sofa_total_48hr,
         sofa_group,
         vent_ever,
         died) %>%
  tbl_summary(
    label = list(age_years = "Age",
                 sofa_total_48hr = "48-hour Maximum SOFA score at beginning of LSE",
                 sofa_group = "SOFA distribution",
                 vent_ever = "Ever on mechanical ventilation",
                 race = "Race/ethnic group",
                 sex = "Sex",
                 died = "Death or discharge to hospice"),
    statistic = age_years ~ "{mean} ({sd})"
  )

tbl1_NW_lse

save(tbl1_NW_lse, file="table1_NW_lse.RData")

gt::gtsave(as_gt(tbl1_NW_lse), filename = "table1_NW_lse.png")
```

## View mortality by age and SOFA group

``` {r age_groups, echo=FALSE, include=FALSE}

age_groups <- data %>%
  ungroup() %>%
  select(age_years, died, sofa_total_48hr, sofa_group) %>%
  mutate(decile = cut(x=age_years,
                      breaks=c(-Inf, 40, 50, 60, 70, 80, Inf)))

age_group1 <- age_groups %>%
  filter(sofa_total_48hr < 8) %>%
  group_by(decile) %>%
  summarise(mortality = mean(died))
age_group1_counts <- age_groups %>%
  filter(sofa_total_48hr < 8) %>%
  pull(decile) %>%
  table()
age_group1$count <- as.numeric(age_group1_counts)

age_group2 <- age_groups %>%
  filter(sofa_total_48hr >= 8 & sofa_total_48hr <= 11) %>%
  group_by(decile) %>%
  summarise(mortality = mean(died))
age_group2_counts <- age_groups %>%
  filter(sofa_total_48hr >= 8 & sofa_total_48hr <= 11) %>%
  pull(decile) %>%
  table()
age_group2$count <- as.numeric(age_group2_counts)

age_group3 <- age_groups %>%
  filter(sofa_total_48hr > 11) %>%
  group_by(decile) %>%
  summarise(mortality = mean(died))
age_group3_counts <- age_groups %>%
  filter(sofa_total_48hr > 11) %>%
  pull(decile) %>%
  table()
age_group3$count <- as.numeric(age_group3_counts)

age_group_all <- age_groups %>%
  group_by(sofa_group, decile) %>%
  summarise(mortality = mean(died))
age_group_all$count <- age_groups %>%
  group_by(sofa_group, decile) %>%
  count() %>%
  pull(n)
age_group_all <- age_group_all %>%
  mutate(se = sqrt(mortality * (1-mortality) / (count - 1)))

save(age_group_all, file="age_group_all.RData")

```

Mortality by age group, stratified by SOFA (all data)
This combines all SOFA groups into a single plot with 95% CI.
``` {r age_group_all, echo=FALSE}

plt1 <- ggplot(data=age_group_all,
               aes(x=decile, y=mortality, group=sofa_group, color=sofa_group)) +
  geom_point(size=2) +
  geom_errorbar(aes(ymin=mortality-1.96*se, ymax=mortality+1.96*se), width=0.1) +
  geom_line() +
  theme_bw() +
  theme(
    panel.background = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = c(0.9, 0.15),
    legend.title = element_blank(),
    legend.background = element_rect(size=0.5, color="black", fill="gray92"),
    legend.key = element_rect(fill="gray92"),
    legend.spacing.y = unit(0, "cm")
  ) +
  xlab("Age Groups (years)") + 
  ylab("Mortality") +
  scale_x_discrete(labels=c("< 40",
                            "40-50",
                            "50-60",
                            "60-70",
                            "70-80",
                            "> 80")) +
  scale_color_manual(values=c("darkgreen", "darkblue", "darkred"))

plt1

save(plt1, file="age_group_all.RData")

ggsave(filename="age_group_all.png")

```

Heatmap of mortality by SOFA and age (all data)
``` {r heatmap1, echo=FALSE}
# make heatmap with observed mortality by 3 SOFA groups and 6 age groups
# 3x6

# 3 categories of sofa
score_rainbow <- data %>%
  ungroup() %>%
  select(died, age_years, sofa_group) %>%
  mutate(age_group = cut(x=age_years,
                         breaks=c(-Inf, 40, 50, 60, 70, 80, Inf))) %>%
  group_by(age_group, sofa_group) %>%
  summarise(mortality=mean(died))

ggplot(data=score_rainbow, mapping=aes(x=age_group, y=sofa_group)) +
  geom_raster(aes(fill=mortality)) +
  scale_fill_gradientn(colors=rev(c("#67001F","#B2182B","#D6604D","#F4A582","#FDDBC7","#D1E5F0","#92C5DE","#4393C3","#2166AC", "#053061")),
                       limits=c(0,1),
                       name="Mortality",
                       na.value="gray20") +
  theme_bw() +
  theme(panel.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text = element_text(size = 14, color="black"),
        axis.title = element_text(size = 16),
        legend.position = "right",
        #legend.title = element_text("Mortality"),
        legend.text = element_text(size = 8, color="black"),
        legend.direction = "vertical",
        legend.key.width = unit(0.5, 'cm'),
        legend.key.height = unit(0.5, 'cm'),
        legend.spacing.y = unit(0.5, 'cm'),
        legend.key = element_rect(fill="gray92"),
        legend.background = element_rect(size=0.5, color="black", fill="gray92")
  ) +
  xlab("Age Group (years)") +
  ylab("48-hour Maximum SOFA Score") +
  scale_y_discrete(labels=c("0-7", "8-11", "> 11"),
                   expand=c(0,0)) +
  scale_x_discrete(labels=c("< 40",
                            "40-50",
                            "50-60",
                            "60-70",
                            "70-80",
                            "> 80"),
                   expand=c(0,0))

ggsave("heatmap_3x6.png")
save(score_rainbow, file="heatmap_3x6.RData")
```


Heatmap of mortality by SOFA and age (all data)
``` {r heatmap2, echo=FALSE}

# make heatmap with observed mortality by 5 SOFA groups and 6 age groups
# 5x6

# 5 categories (based on quantiles n = 5)
score_rainbow <- data %>%
  ungroup() %>%
  select(died, age_years, sofa_total_48hr) %>%
  mutate(age_group = cut(x=age_years,
                         breaks=c(-Inf, 40, 50, 60, 70, 80, Inf)),
         sofa_group = cut(x=sofa_total_48hr,
                          breaks=c(-Inf, 4, 5, 6, 8, Inf))) %>%
  group_by(age_group, sofa_group) %>%
  summarise(mortality=mean(died))

ggplot(data=score_rainbow, mapping=aes(x=age_group, y=sofa_group)) +
  geom_raster(aes(fill=mortality)) +
  scale_fill_gradientn(colors=rev(c("#67001F","#B2182B","#D6604D","#F4A582","#FDDBC7","#D1E5F0","#92C5DE","#4393C3","#2166AC", "#053061")),
                       limits=c(0,1),
                       name="Mortality",
                       na.value="gray20") +
  theme_bw() +
  theme(panel.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text = element_text(size = 14, color="black"),
        axis.title = element_text(size = 16),
        legend.position = "right",
        #legend.title = element_text("Mortality"),
        legend.text = element_text(size = 8, color="black"),
        legend.direction = "vertical",
        legend.key.width = unit(0.5, 'cm'),
        legend.key.height = unit(0.5, 'cm'),
        legend.spacing.y = unit(0.5, 'cm'),
        legend.key = element_rect(fill="gray92"),
        legend.background = element_rect(size=0.5, color="black", fill="gray92")
  ) +
  xlab("Age Group (years)") +
  ylab("48-hour Maximum SOFA Score") +
  scale_y_discrete(labels=c("< 5", "5", "6", "7-8", "> 8"),
                   expand=c(0,0)) +
  scale_x_discrete(labels=c("< 40",
                            "40-50",
                            "50-60",
                            "60-70",
                            "70-80",
                            "> 80"),
                   expand=c(0,0))

ggsave("heatmap_5x6.png")
save(score_rainbow, file="heatmap_5x6.RData")
```


## Logistic Regression

In this section we will make several LR models:

* SOFA (continuous variable)
* SOFA categories (NY guideline cutoffs)
* SOFA + age (continuous variables)

```{r make_cats, include=FALSE}
# make factor of SOFA categories based off NY guidelines
data <- data %>%
  mutate(sofa_cat = case_when(
    sofa_total_48hr < 8 ~ "SOFA <= 7",
    sofa_total_48hr < 12 ~ "SOFA 8-11",
    sofa_total_48hr >= 12 ~ "SOFA >= 12"
  ))
```

The LR model data will be saved and sent to the "test" data institution.
``` {r make_LRs, include=FALSE}

# Change outcome to factor.
data <- data %>%
  mutate(died = ifelse(died==0, "No", "Yes"),
         died = as.factor(died))
data$died <- relevel(data$died, ref="No")

# create LRs
lr_sofa <- glm(died ~ sofa_total_48hr,
               data=data,
               family="binomial")

lr_sofa_cat <- glm(died ~ sofa_cat,
                   data=data,
                   family="binomial")

lr_sofa_age <- glm(died ~ sofa_total_48hr + age_years,
                   data=data,
                   family="binomial")

save(lr_sofa, lr_sofa_cat, lr_sofa_age, file="LR_models.RData")

```

## Create Risk Score

``` {r risk_score}

# from lr_sofa_age

# Example:
# (note: these numbers are from when model was run on entire UC dataset)
# intercept estimate: -4.56
# sofa estimate: 0.212811 --> OR 1.23
# age estimate: 0.041470 --> OR 1.04

# sofa range 0 - 24
# age range 18 - 100

# 1 pt sofa about 5x as bad as 1 year
# each year increase 1/5 pt
# each sofa increase 1 pt

# round to nearest whole

#r = 0.2*(age-18) + SOFA

# The code below will extract these numbers and automate it, but it should be similar.
coeff_sofa <- lr_sofa_age$coefficients[2]
coeff_age <- lr_sofa_age$coefficients[3]
m <- round(coeff_age / coeff_sofa, digits=2) # multiplier

# risk score r = m * (age - 18) + SOFA

data <- data %>%
  mutate(score = m*(age_years - 18) + sofa_total_48hr,
         score = ifelse(age_years > 98, m*80 + sofa_total_48hr, score))

roc_score_train <- roc(response=data$died,
                  predictor=data$score,
                  auc=TRUE)

lr_score <- glm(died ~ score,
                data=data,
                family="binomial")

data$pred_score <- predict(lr_score, data)
data <- data %>%
  mutate(prob_score = exp(pred_score)/(1+exp(pred_score)))

#training AUC
roc_score_train$auc

# save model
save(lr_score, roc_score_train, m, file="risk_score.RData")
```

``` {r make_preds, echo=FALSE}
# make predictions for calibration plot creation
data$pred_sofa <- predict(lr_sofa, data)
data <- data %>%
  mutate(prob_sofa = exp(pred_sofa)/(1+exp(pred_sofa)))
```

``` {r cal_byage_plot, echo=FALSE}

# Make plot of observed mortality and predicted mortality for age groups.

# change died back to 0/1
data$died <- ifelse(data$died == "Yes", 1, 0)

age_groups <- data %>%
  ungroup() %>%
  select(age_years, died, prob_sofa, prob_score) %>%
  mutate(age_group = cut(x=age_years,
                      breaks=c(-Inf, 40, 50, 60, 70, 80, Inf))) %>%
  select(-age_years)

age_groups_sum <- age_groups %>%
  group_by(age_group) %>%
  summarise(mortality = mean(died)) %>%
  mutate(group = "Observed Mortality")

age_groups_sum <- age_groups %>%
  group_by(age_group) %>%
  summarise(mortality = mean(prob_score)) %>%
  mutate(group = "Predicted Mortality by APSI Score") %>%
  rbind(age_groups_sum)

age_groups_sum <- age_groups %>%
  group_by(age_group) %>%
  summarise(mortality = mean(prob_sofa)) %>%
  mutate(group = "Predicted Mortality by SOFA Score") %>%
  rbind(age_groups_sum)

ggplot(data=age_groups_sum, aes(x=age_group,
                                y=mortality,
                                fill=group)) +
  geom_col(position="dodge",
           color="black") +
  scale_fill_manual(values=c("gray30", "dodgerblue", "indianred")) +
  xlab("Age Group (years)") +
  ylab("Life-support Episode Mortality (%)") +
  theme_bw() +
  theme(panel.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text = element_text(size = 14, color="black"),
        axis.title = element_text(size = 16),
        legend.position = c(0.25,0.8),
        legend.title = element_blank(),
        legend.text = element_text(size = 8, color="black"),
        legend.direction = "vertical",
        legend.key.width = unit(0.5, 'cm'),
        legend.key.height = unit(0.5, 'cm'),
        legend.spacing.y = unit(0, 'cm'),
        legend.key = element_rect(fill="gray92"),
        legend.background = element_rect(size=0.5, color="black", fill="gray92")
  ) +
  scale_x_discrete(labels=c("< 40",
                            "40-50",
                            "50-60",
                            "60-70",
                            "70-80",
                            "> 80")) +
  scale_y_continuous(breaks=seq(0.1, 0.9, 0.1),
                     labels=c("10", "20", "30", "40", "50", "60", "70", "80", "90"),
                     limits=c(0, 0.7))# +
  #geom_text(data=score_data_tiled_gg, aes(label=mortality))

ggsave(filename="Calibration_bar_byage_UCNW.png")
save(age_groups, file="age_groups_predictions_UCNW.RData")

```

``` {r cal_byscore_plot, echo=FALSE}

# Make plot of observed mortality and predicted mortality by deciles of each score (SOFA or APSI).

score_data <- data %>%
  ungroup() %>%
  select(sofa_total_48hr, score, prob_sofa, prob_score, died) %>%
  mutate(sofa_deciles = ntile(sofa_total_48hr, 10),
         score_deciles = ntile(score, 10))

score_data_tiled <- score_data %>%
  group_by(sofa_deciles) %>%
  summarise(obs_sofa_mortality = mean(died)) %>%
  rename(deciles = sofa_deciles)

score_data_tiled <- score_data %>%
  group_by(score_deciles) %>%
  summarise(obs_score_mortality = mean(died)) %>%
  rename(deciles = score_deciles) %>%
  full_join(score_data_tiled, by="deciles")

score_data_tiled <- score_data %>%
  group_by(score_deciles) %>%
  summarise(pred_score_mortality = mean(prob_score)) %>%
  rename(deciles = score_deciles) %>%
  full_join(score_data_tiled, by="deciles")

score_data_tiled <- score_data %>%
  group_by(score_deciles) %>%
  summarise(pred_sofa_mortality = mean(prob_sofa)) %>%
  rename(deciles = score_deciles) %>%
  full_join(score_data_tiled, by="deciles")

score_data_tiled <- score_data %>%
  group_by(sofa_deciles) %>%
  summarise(pred_sofa_mortality_by_sofa_deciles = mean(prob_sofa)) %>%
  rename(deciles = sofa_deciles) %>%
  full_join(score_data_tiled, by="deciles")

score_data_tiled_gg <- score_data_tiled %>%
  select(deciles, obs_score_mortality) %>%
  mutate(group = "Observed Mortality") %>%
  rename(mortality = obs_score_mortality)

score_data_tiled_gg <- score_data_tiled %>%
  select(deciles, pred_score_mortality) %>%
  mutate(group = "Predicted Mortality by APSI Score") %>%
  rename(mortality = pred_score_mortality) %>%
  rbind(score_data_tiled_gg)

score_data_tiled_gg <- score_data_tiled %>%
  select(deciles, pred_sofa_mortality) %>%
  mutate(group = "Predicted Mortality by SOFA Score") %>%
  rename(mortality = pred_sofa_mortality) %>%
  rbind(score_data_tiled_gg)

ggplot(data=score_data_tiled_gg, aes(x=deciles,
                                     y=mortality,
                                     fill=group)) +
  geom_col(position="dodge",
           color="black") +
  scale_fill_manual(values=c("gray30", "dodgerblue", "indianred")) +
  xlab("Deciles of Risk or SOFA Score") +
  ylab("Life-support Episode Mortality (%)") +
  theme_bw() +
  theme(panel.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text = element_text(size = 14, color="black"),
        axis.title = element_text(size = 16),
        legend.position = c(0.25,0.8),
        legend.title = element_blank(),
        legend.text = element_text(size = 8, color="black"),
        legend.direction = "vertical",
        legend.key.width = unit(0.5, 'cm'),
        legend.key.height = unit(0.5, 'cm'),
        legend.spacing.y = unit(0, 'cm'),
        legend.key = element_rect(fill="gray92"),
        legend.background = element_rect(size=0.5, color="black", fill="gray92")
  ) +
  scale_x_continuous(breaks=c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10),
                   labels=c("1", "2", "3", "4", "5", "6", "7", "8", "9", "10")) +
  scale_y_continuous(breaks=seq(0.1, 0.9, 0.1),
                     labels=c("10", "20", "30", "40", "50", "60", "70", "80", "90"))# +
  #geom_text(data=score_data_tiled_gg, aes(label=mortality))

ggsave(filename="Calibration_bar_byscore_UCNW.png")
save(score_data_tiled, file="score_data_tiled_UCNW.RData")
  
```

``` {r mortality_by_race}

mort_race <- data %>%
  group_by(race) %>%
  summarise(mortality = mean(prob_sofa)) %>%
  as.data.frame() %>%
  mutate(group = "Predicted Mortality by SOFA Score")

mort_race <- data %>%
  group_by(race) %>%
  summarise(mortality = mean(prob_score)) %>%
  as.data.frame() %>%
  mutate(group = "Predicted Mortality by APSI Score") %>%
  rbind(mort_race)

mort_race <- data %>%
  group_by(race) %>%
  summarise(mortality = mean(died)) %>%
  as.data.frame() %>%
  mutate(group = "Observed Mortality") %>%
  rbind(mort_race)

mort_race$race <- factor(mort_race$race, levels=c("Non-Hispanic White", "Non-Hispanic Black", "Hispanic", "Other"))

ggplot(data=mort_race, aes(x=race,
                           y=mortality,
                           fill=group)) +
  geom_col(position="dodge",
           color="black") +
  scale_fill_manual(values=c("gray30", "dodgerblue", "indianred")) +
  xlab("Race/Ethnicity") +
  ylab("Life-support Episode Mortality (%)") +
  theme_bw() +
  theme(panel.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text = element_text(size = 14, color="black"),
        axis.title = element_text(size = 16),
        legend.position = c(0.3,0.85),
        legend.title = element_blank(),
        legend.text = element_text(size = 8, color="black"),
        legend.direction = "vertical",
        legend.key.width = unit(0.5, 'cm'),
        legend.key.height = unit(0.5, 'cm'),
        legend.spacing.y = unit(0, 'cm'),
        legend.key = element_rect(fill="gray92"),
        legend.background = element_rect(size=0.5, color="black", fill="gray92")
  ) +
  scale_x_discrete(labels=c("Non-Hispanic\nWhite",
                            "Non-Hispanic\nBlack",
                            "Hispanic",
                            "Other")) +
  scale_y_continuous(breaks=seq(0, .4, 0.1),
                     labels=c("0", "10", "20", "30", "40"),
                     limits=c(0, 0.4))# +
  #geom_text(data=score_data_tiled_gg, aes(label=mortality))

ggsave(filename="Calibration_bar_byrace_UCNW.png")

save(mort_race, file="Mortality_by_race_UCNW.RData")
```

``` {r statistics_race}
data %>% distinct(patient, .keep_all=TRUE) %>%
  group_by(race) %>%
  summarise(median_age = median(age_years),
            iqr_age = IQR(age_years),
            mean_age = mean(age_years),
            sd_age = sd(age_years))
```

``` {r statistics_race_sofa}
data %>% group_by(race) %>%
  summarise(sofa = median(sofa_total_48hr),
            iqr_sofa = IQR(sofa_total_48hr))
```