---
title: "Quality control checks on data for age and life-support triage project"
output:
  html_document:
    toc: yes
    toc_depth: '2'
    df_print: paged
  pdf_document: default
  html_notebook:
    theme: journal
    toc: yes
    toc_depth: 2
    toc_float: yes
---

```{r packages, results=FALSE, message=FALSE, warning=FALSE}
library(dplyr)
library(gtsummary)
library(pROC)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

This script accepts a clean data file which has been appropriately formatted for this age and life-support triage project (see url: https://github.com/08wparker/age_life_support_triage) and performs various quality control checks on it.

## Load data in and rearrange columns

```{r data_in}

data <- read.csv("C:/Users/t.cri.mhermsen/Documents/UCMC_data_clean/WashU/bjc_deidentified_2023-01-28.csv") %>%
  rename(died = dead)
#data <- read.csv("C:/Users/t.cri.mhermsen/Documents/UCMC_data_clean/Run_UC_NW_train/data_for_analysis.csv")

data <- data %>%
  arrange(patient, encounter, lfspprt_episode)

```

``` {r sofagroups, echo=FALSE}
# create SOFA groups variable for table 1s
data <- data %>%
  group_by(encounter, lfspprt_episode) %>%
  mutate(sofa_group = cut(sofa_total_48hr, breaks=c(-Inf, 7, 11, Inf),
                          labels= c("SOFA 0-7",
                                    "SOFA 8-11",
                                    "SOFA > 11")))

```

# Initial sample statistics

Create some tables of patient-level, encounter-level, and LSE-level statistics.

First, patient-level demographics. Note that race/ethnicity should be reported as a single variable with the following levels: Non-Hispanic White, Non-Hispanic Black, Hispanic, Other.
```{r table1_pt, echo=FALSE}

tbl1_pt <- data %>%
  ungroup %>%
  arrange(patient, encounter, lfspprt_episode) %>%
  distinct(patient, .keep_all=TRUE) %>% # keeps first instance
  select(age_years,
         sex,
         race,
         died) %>%
  tbl_summary(
    label = list(age_years = "Age",
                 race = "Race/ethnic group",
                 sex = "Sex",
                 died = "Death or discharge to hospice"),
    statistic = age_years ~ "{mean} ({sd})"
  )

tbl1_pt

```

Histogram of patient ages:

``` {r ages_hist, echo=FALSE}
data %>%
  ungroup() %>%
  select(patient, age_years) %>%
  distinct(patient, .keep_all=TRUE) %>%
  pull(age_years) %>%
  hist(main="Histogram of Patient Ages (in years)")

```

Youngest patient (should be 18 or older): `r min(data$age_years, na.rm=TRUE)`

Oldest patient: `r max(data$age_years, na.rm=TRUE)`

Encounter-level statistics:
```{r table1_enc, echo=FALSE}

tbl1_enc <- data %>%
  ungroup %>%
  arrange(patient, encounter, lfspprt_episode) %>%
  distinct(encounter, .keep_all=TRUE) %>% # keeps first instance
  select(died) %>%
  tbl_summary(
    label = list(died = "Death or discharge to hospice")
  )

tbl1_enc

```

Now, LSE-level statistics:
```{r table1_lse, echo=FALSE}

tbl1_lse <- data %>%
  ungroup %>%
  arrange(patient, encounter, lfspprt_episode) %>% # no filtering to do, each row is unique LSE
  select(sofa_total_48hr,
         sofa_group,
         vent_ever,
         died) %>%
  tbl_summary(
    label = list(sofa_total_48hr = "48-hour Maximum SOFA score at beginning of LSE",
                 sofa_group = "SOFA distribution",
                 vent_ever = "Ever on mechanical ventilation",
                 died = "Death or discharge to hospice")
  )

tbl1_lse
```

Note: the "died" variable should be reported at the level of the LSE. That is, "died" should be 1 if the patient died at the end of the current LSE or 0 if the patient lived. Since each row corresponds to a unique LSE and patients may have more than one LSE, the variable "died" should take value 1 for exactly 1 row per patient.

``` {r died, echo=FALSE}
times_died <- data %>%
  group_by(patient) %>%
  summarise(times_died = sum(died)) %>%
  pull(times_died) %>%
  max(na.rm=TRUE)
```

In other words, each patient should die only once... In this dataset, that maximum number of times a patient died was `r times_died` times (again, should be 1).

## Examine number of patients and LSEs
In the training data, there were approximately 3,611 patients corresponding to 5,648 LSEs (1.56 LSEs per patient, median 1 LSE per patient). The highest number of LSEs for a single patient was 18.
``` {r pt_lse_comparison, echo=FALSE}
num_pts <- data %>%
  ungroup %>%
  select(patient) %>%
  unique() %>%
  nrow()

# if appropriately coded, data should already be 1 row per LSE
num_lse <- data %>%
  nrow()

summary_lse <- data %>%
  group_by(encounter) %>%
  filter(lfspprt_episode == max(lfspprt_episode)) %>%
  pull(lfspprt_episode) %>%
  summary()

```

This dataset has `r num_pts` patients and `r num_lse` LSEs. There are approximately `r round(num_lse/num_pts,2)` LSEs per patient and a median of `r summary_lse["Median"]` LSEs. At least 1 patient in this dataset had `r summary_lse["Max."]` LSEs.

Histograms of number of LSEs:

``` {r lse_hist, echo=FALSE}
data %>%
  group_by(encounter) %>%
  filter(lfspprt_episode == max(lfspprt_episode)) %>%
  pull(lfspprt_episode) %>%
  hist(main="Histogram of Number of LSEs (untrimmed)")
```
``` {r lse_hist_trimmed, echo=FALSE}
data %>%
  group_by(encounter) %>%
  filter(lfspprt_episode == max(lfspprt_episode)) %>%
  pull(lfspprt_episode) %>%
  hist(main="Histogram of Number of LSEs (trimmed for viewing)",
       xlim=c(0,30),
       ylim=c(0, round(num_pts*0.75, 0)),
       freq=TRUE)
```

## Check Proportion of COVID positive
Datasets should be filtered so that LSEs are coming from 90% COVID+ encounters and 10% COVID- encounters. See Github README (url in introduction) for details on how COVID positivity is defined.

``` {r covid, echo=FALSE}
covid_freq <- data %>%
  ungroup() %>%
  distinct(encounter, .keep_all=TRUE) %>%
  pull(covid) %>%
  mean()
```

In this dataset, `r 100*covid_freq`% of your encounters were COVID+.

## Examine SOFA Scores
Discrete distributions of SOFA scores based on 3 categories (< 7, 8-11, > 11) can be found in the LSE-level Table 1 generated above.

Histogram of 48-hour maximum SOFA scores at the beginning of LSE:

``` {r sofa_hist, echo=FALSE}
data %>%
  ungroup() %>%
  pull(sofa_total_48hr) %>%
  hist(main="Histogram of 48-hour Maximum SOFA score at beginning of LSE")
```

Ideally, this histogram will show steadily decreasing numbers of higher SOFA scores (approximately geometric distribution).

Lowest SOFA score (should be 0, 1, or 2): `r min(data$sofa_total_48hr, na.rm=TRUE)`

Highest SOFA score (should be 24 or less): `r max(data$sofa_total_48hr, na.rm=TRUE)`

Finally, we will examine how well SOFA alone predicts mortality. If coded accurately, it should be somewhat able to predict ICU mortality (though we aim to show this ability is much improved if age is also considered).

For reference, 48-hour maximum SOFA score at the beginning of the LSE predicted LSE mortality with an AUC of 0.62.

``` {r sofa_predict}
lr <- glm(died ~ sofa_total_48hr,
          data=data,
          family="binomial")

roc_score <- roc(response=data$died,
                 predictor=lr$fitted.values,
                 auc=TRUE)
```

In this dataset, 48-hour maximum SOFA score at the beginning of the LSE predicts LSE mortality with an AUC of `r roc_score$auc`.

## Examine SOFA Sub-Scores

Note: each of these 48-hour maximum sub-scores are at the beginning of the specified LSE.

This section requires that data contain columns for each 48-hour maximum sub-score, which may not have been supplied by other institutions. If the data does not contain these columns (such as "sofa_cv_48hr"), then do not run the chunks in this section.

``` {r sofa_cv_hist, echo=FALSE, eval=FALSE}
data %>%
  ungroup() %>%
  pull(sofa_cv_48hr) %>%
  hist(main="Histogram of 48-hour Maximum SOFA Cardiovascular Sub-score")
```

``` {r sofa_resp_hist, echo=FALSE, eval=FALSE}
data %>%
  ungroup() %>%
  pull(sofa_resp_48hr) %>%
  hist(main="Histogram of 48-hour Maximum SOFA Respiratory Sub-score")
```

``` {r sofa_renal_hist, echo=FALSE, eval=FALSE}
data %>%
  ungroup() %>%
  pull(sofa_renal_48hr) %>%
  hist(main="Histogram of 48-hour Maximum SOFA Renal Sub-score")
```

``` {r sofa_liver_hist, echo=FALSE, eval=FALSE}
data %>%
  ungroup() %>%
  pull(sofa_liver_48hr) %>%
  hist(main="Histogram of 48-hour Maximum SOFA Liver Sub-score")
```

``` {r sofa_coags_hist, echo=FALSE, eval=FALSE}
data %>%
  ungroup() %>%
  pull(sofa_coags_48hr) %>%
  hist(main="Histogram of 48-hour Maximum SOFA Coagulation Sub-score")
```

``` {r sofa_neuro_hist, echo=FALSE, eval=FALSE}
data %>%
  ungroup() %>%
  pull(sofa_neuro_48hr) %>%
  hist(main="Histogram of 48-hour Maximum SOFA Neurologic Sub-score")
```